\chapter{Conclusiones}
	
	
	
	
	
	
	\section{Fichas bibliográficas}


		\begin{enumerate}
				\item 2021 - Luis Gerardo De La Fraga - Designing an authenticated Hash function with a 2D chaotic map. \cite{Fraga2021}
				\item 2017 - Luis Gerardo De La Fraga - Hardware implementation of pseudo-random number generators based on chaotic maps. \cite{Fraga2017}
				\item 2021 - Sundarapandian Vaidyanathan - A 5-D Multi-Stable Hyperchaotic Two-Disk Dynamo System With No Equilibrium Point: Circuit Design, FPGA Realization and Applications to TRNGs and Image Encryption. \cite{Vaidyanathan2021}
				\item 2020 - N. Nalla Anandakumar - FPGA-Based True Random Number Generation Using Programmable Delays in Oscillator-Rings. \cite{Anandakumar2020}
				\item 1999 - Benjamin Jun and Paul Kocher - The Intel Random Number Generator. \cite{Jun1999}
				\item 2021 - Xinyu Wang - High-Throughput Portable True Random Number Generator Based on Jitter-Latch Structure. \cite{Wang2021}
				\item 2021 - Adriaan Peetermans - Design and Analysis of Configurable Ring Oscillators for True Random Number Generation Based on Coherent Sampling. \cite{Peetermans2021}
				\item 2016 - Oto Petura - A Survey of AIS-20/31 Compliant TRNG Cores Suitable for FPGA Devices. \cite{Petura2016}
				\item 1993 - J.C. Sprott - Automatic generation of strange attractors. \cite{Sprott1993}
				\item 2015 - Patrick Haddad - A Physical Approach for Stochastic Modeling of TERO-Based TRNG. \cite{Haddad2015}
				\item 2010 - Mathieu Baudet - On the Security of Oscillator-Based Random Number Generators.  \cite{Baudet2010}
				\item 2010 - Michal Varchola - New High Entropy Element for FPGA Based True Random Number Generators. \cite{Varchola2010}
				\item 2019 - Viktor Fischer -  Modern random number generator design – Case study on a secured PLL-based TRNG \cite{Fischer2019}
				\item 2021 - Soyeon Choi - Analysis of Ring Oscillator based True Random Number Generator on FPGAs  \cite{Choi2021}
				\item 2021 - Claudio Garcia - Chaotic Features of a Class of Discrete Maps without Fixed Points  \cite{GarciaGrimaldo2021}
				\item 2020 - Rethinam Sivaraman - Ring oscillator as confusion diffusion agent: a complete TRNG drove image security  \cite{Sivaraman2020}
				\item 2020 - Joan Daemen - The design of Rijndael: AES The Advanced Encryptation Standard.\cite{JoanDaemen2020}
				\item 2003 - Viktor Fischer - True Random Number Generator Embedded in Reconfigurable Hardware \cite{Fischer2003}	
				\item 2018 - Viktor Fischer - From Physical to Stochastic Modeling of a TERO-Based TRNG  \cite{Bernard2018}	
				\item 2011 - Viktor Fischer - Security Trends for FPGAS   \cite{Badrignans2011}
				
				\item 2016 - Viktor Fischer - Platform for Testing and Evaluation of PUF and TRNG Implementations in FPGAs   \cite{Laban2016}
			\end{enumerate}
	